---
title: "SelectProbesBySimulation"
author: "Nathan Gerald Skene"
date: "11 July 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the scripts

```{r }
source("R/SelectGeneSet.r")
source("R/GetGeneSetOptions.r")
source("R/GetLowerAnnotGroups.r")
source("R/CheckGeneSetProportions.r")
source("R/GetTrainingAndTestSets.r")
source("R/GetPredictionError.r")
source("R/plotOptI.r")
source("R/plotGeneSet.r")
#source("GeneSetSummary.r")
source("R/SetLevels.r")
source("R/LoadExpData.r")
source("R/GetMeans.r")
source("R/SelectCandidates.r")
source("R/SelectCandidatesForPairs.r")
```

## Prepare the data and celltype level annotations

We will use the read count expression data, with multiples cells per class. And then annotated these with multiple levels.

The data is structured such that $cellNames is an array of cell type names, equal in length to the number of columns in $readCounts.

```{r }
# Load the read counts
expDat = LoadExpData()

# Check the data
print(expDat$cellNames[1:3])
print(expDat$readCounts[1:4,1:6])
```

The manner in which probes are selected assumes that you have mapped out a heirarchical tree across which the cell types are distibuted, i.e. Int16 is a subtype of interneuron, which is a subtype of neuron.

The SetLevels() function is specifically designed for this dataset. It can have an arbitrary number of levels. 

levels[[1]]$annot would have an entry for each cell in expData$readCounts, the entry for each cell would be something like either 'Neurons' or 'Glia'.

Some cell types may be divided into many layers, others might only be considered 'glia' and not subdivided further. In which case, at higher levels just enter them as '-'.

```{r }
# Setup level annotations
levels = SetLevels(expDat$cellNames)
```

## Find means & proportions for each level

To find candidate genes for each level, we will find the mean level of expression within each group, and the proportion of expression within that group (relative only to the cells in the level below).

```{r }
for(targLevel in 1:length(levels)){
#for(targLevel in 4){
    # Get the means, proportions etc
    levels[[targLevel]]$stats = GetMeans(expDat,annot=levels[[targLevel]]$annot)
    
    # Select candidate genes
    count=0
    for(target in setdiff(unique(levels[[targLevel]]$annot),"-")){
        count=count+1
        
        # Enforce constraints on candidates
        gMeans = levels[[targLevel]]$stats$groupMeans#groupMedians#$groupMeans
        props  = levels[[targLevel]]$stats$proportions
        a      = levels[[targLevel]]$annot
        if(targLevel==1){a0 = NA}else{a0 = levels[[targLevel-1]]$annot}
        # target=target
        # groupMeans=gMeans
        # proportions=props
        # annot=a
        # annotBelow=a0
        # numCandidates=40
        #candidates = SelectCandidates(target=target,groupMeans=gMeans,proportions=props,annot=a,annotBelow=a0,numCandidates=20)
        candidates = SelectCandidatesForPairs(target=target,groupMeans=gMeans,proportions=props,annot=a,annotBelow=a0,numCandidates=3)
        print(candidates)
        
        if(!is.na(candidates)){
            if(count==1){allCandidates=candidates}
            else{allCandidates=rbind(allCandidates,candidates)}
        }
    }
    levels[[targLevel]]$candidates=allCandidates
}
```

# Get a table with all candidates
```{r }
allCands = levels[[1]]$candidates
for(i in 2:6){
    allCands = rbind(allCands,levels[[i]]$candidates)
}
write.csv(allCands,file="allCands.csv")
```

## I manually annotated allCands.csv with a "Decision" column...
## Then Kenneth and I discussed the output of the prior linear model method..
## Then kenneth added a few extra
```{r }
myProbes = read.csv("/Users/ns9/Google Drive/In Situ Probe Selection/allCands_manuallyAnnotated.csv",stringsAsFactors = FALSE)
myProbes = myProbes[myProbes$Decision=="Buy",]
disProbes = read.csv("/Users/ns9/Google Drive/In Situ Probe Selection/Post_Discussion_Probelist.csv",stringsAsFactors = FALSE)
disProbes = disProbes[disProbes$Decision=="Buy",]

howManyGenes = unique(c(myProbes$genes,disProbes$symbols))

# How many of the genes are:
# - In the new gene list
# - Not in the triplet gene list
# - Not previously suggested by Kenneth
NotInTriplet = setdiff(myProbes$genes,unique(disProbes$symbols))
NotSuggested = NotInTriplet[!NotInTriplet %in% kenneths$GENE]
myProbes[myProbes$genes %in% NotSuggested,]

as.character(kenneths$GENE)
```


## For each level, find each combination of markers that could be used

```{r }
save(levels,expDat,file=sprintf("dataForServer-%s.rda",gsub(':|-', '_', Sys.time())))
library(combinat)
library(reshape2)
numPerLevel=3
count=0
for(targLevel in 1:length(levels)){
    
    # For each "group" find the three genes which perform best
    for(g in unique(levels[[targLevel]]$annot)){
        count=count+1
        print(sprintf("%s - %s - %s",count,targLevel,g))
        #print(g)
    }
}
        
        output = SelectGeneSet(target=g,targLevel=targLevel,levels=levels,readCounts=expDat$readCounts)
        
        if(g==unique(levels[[targLevel]]$annot)[1]){
            levels[[targLevel]]$selectedMarkers=output$selectedMarkers
            levels[[targLevel]]$marker_stats=output$GeneSetSummary
        }else{
            levels[[targLevel]]$selectedMarkers=rbind(levels[[targLevel]]$selectedMarkers,output$selectedMarkers)
            levels[[targLevel]]$marker_stats=rbind(levels[[targLevel]]$marker_stats,output$GeneSetSummary)
        }
    }
}

save(levels,file=sprintf("levels-%s.rda",gsub(':|-', '_', Sys.time())))
```

## Script for the server

Created a folder at /nfs/team156/ns9/In Situ Probe Selection/

Put below script in 'runSelection.r'

```{r }
load(file="dataForServer-2016_07_15 14_13_54.rda")
library(combinat)
library(reshape2)
library(ggplot2)
numPerLevel=3

source("SelectGeneSet.r")
source("GetGeneSetOptions.r")
source("GetLowerAnnotGroups.r")
source("CheckGeneSetProportions.r")
source("GetTrainingAndTestSets.r")
source("GetPredictionError.r")
source("plotOptI.r")
source("GeneSetSummary.r")
#source(".r")


for(targLevel in 1:length(levels)){
    
    # For each "group" find the three genes which perform best
    for(g in unique(levels[[targLevel]]$annot)){
         output = SelectGeneSet(target=g,targLevel=targLevel,levels=levels,readCounts=expDat$readCounts)
         save(output,file=sprintf("%s---%s.rda",targLevel,g))
    }
}

```

The selection of the final set of genes does not weight by expression level.

```{r }
targLevel=2
target="Pyramidal"

load(sprintf("/Users/ns9/Google Drive/In Situ Probe Selection/Server Output/%s---%s.rda",targLevel,target))
output$GeneSetSummary[1:10,]
newOut=data.frame()
for(i in 1:50){
    rowGenes = as.character(unlist(output$GeneSetSummary[i,2:4]))
    expLevel = levels[[targLevel]]$stats$groupMedians[rowGenes,target]
    tmp = cbind(output$GeneSetSummary[i,],exp1=expLevel[1],exp2=expLevel[2],exp3=expLevel[3])
    newOut=rbind(newOut,tmp)
}
```

Load markers we already have

```{r }
curMarkers = read.csv(file="/Users/ns9/Google Drive/In Situ Analysis/Analysis of probe results in cells.csv")
currently_used_Genes = as.character(curMarkers$genes)
```

Get list of all selected markers:

```{r }
markerSet = data.frame()
originalMarkerSet = data.frame()
for(targLevel in 1:length(levels)){
    targets = as.character(unique(levels[[targLevel]]$candidates[,1]))
    for(target in targets){
        load(sprintf("/Users/ns9/Google Drive/In Situ Probe Selection/Server Output v2/%s---%s.rda",targLevel,target))    
        
        # If any bad genes got through, drop them
        bad_genes = "Xist"
        output$GeneSetSummary = output$GeneSetSummary[!output$GeneSetSummary$g1 %in% bad_genes,]
        output$GeneSetSummary = output$GeneSetSummary[!output$GeneSetSummary$g2 %in% bad_genes,]
        output$GeneSetSummary = output$GeneSetSummary[!output$GeneSetSummary$g3 %in% bad_genes,]
        
        # Search for the highest expressed amongst the top 200th
        # - numEval is the number of top options to try (top 200th)
        numEval = round(dim(output$GeneSetSummary)[1]/200)       
        if(numEval<1){numEval=1}
        # - Instead of arbitary cutoff, expand numEval if neighbouring samples also have equivilent FPos+FNeg
        numEval = sum(apply(output$GeneSetSummary[,c("FPos","FNeg")],1,sum)<=sum(output$GeneSetSummary[numEval,c("FPos","FNeg")]))
        mGenes = output$GeneSetSummary[1:numEval,2:4]
        possMarkers = unique(as.character(unlist(mGenes)))
        mExp = matrix(0,dim(mGenes)[1],dim(mGenes)[2])
        for(i in 1:dim(mGenes)[1]){mExp[i,]=levels[[targLevel]]$stats$groupMedians[as.character(unlist(mGenes[i,])),target]}
        minExp = apply(mExp,1,min)
        useCands = 1
        for(i in 1:dim(mGenes)[1]){
            if(minExp[useCands]<10){
                if(minExp[useCands]<minExp[i]){useCands=i}
            }
        }
        
        print(sprintf("Level %s - %s",targLevel,target))
        print(output$GeneSetSummary[c(1,useCands),c("FPos","FNeg")])
        
        if(target=="Htr3a Vip+ - Int10"){store_tmp = originalMarkerSet}
        
        # Check if any of the top candidate genesets are ones we already have
        numCands = rep(0,numEval)
        for(jj in 1:numEval){
            cands = as.character(unlist(output$GeneSetSummary[jj,2:4]))
            curGenes = unique(c(currently_used_Genes,as.character(originalMarkerSet$symbols)))
            numCands[jj] = sum(cands %in% curGenes)
        }
        if(sum(numCands)>=1){
            #useWhich = which.max(minExp[numCands==max(numCands)])
            useWhich = which.max(minExp[numCands>0])
            useCands=which(numCands>0)[useWhich]
        }
        
        print(useCands)
        originalThree = as.character(unlist(output$GeneSetSummary[useCands,2:4]))
        
        # Force usage of specific marker sets
        if(target=="Interneurons"){originalThree = c("Gad1","Cnr1","Slc6a1")}
        if(target=="Htr3a Vip+ - Int10"){originalThree = c("Crh","Penk","Klhl13")}
        
        # Plot the marker set
        annotGroups=GetLowerAnnotGroups(target,targLevel,levels)
        rC=expDat$readCounts
        annot=levels[[targLevel]]$annot
        a0=annotGroups$annotBelow
        pdf(file=sprintf("FinalMarkerPlots/MarkerPlot-Level%s-%s.pdf",targLevel,target),width=8,height=4)
        plotGeneSet(geneSet=originalThree,target=target,readCounts=rC,annot=annot,annotBelow=a0)
        dev.off()
        
        g_exp = levels[[targLevel]]$stats$groupMedians[originalThree,target]
        # addThree = originalThree
        # addThree = setdiff(addThree,markerSet$symbols)
        # j=1
        # while(length(addThree)<3){
        #     j=j+1
        #     nextThree = as.character(unlist(output$GeneSetSummary[j,2:4]))
        #     nextThree = setdiff(nextThree,markerSet$symbols)
        #     originalThree = c(originalThree,intersect(nextThree,markerSet$symbols))
        #     nextThree = setdiff(nextThree,addThree)
        #     if(length(nextThree)>1){
        #         expLevels = levels[[targLevel]]$stats$groupMedians[nextThree,target]
        #         mostExpressed = names(sort(expLevels,decreasing=TRUE)[1])
        #         addThree = c(addThree,mostExpressed)
        #     }
        #     if(length(nextThree)==1){
        #         addThree = c(addThree,nextThree)
        #     }
        # }
        # addThree = addThree[!is.na(addThree)]
        # allMarkers = unique(c(originalThree,addThree))
        # tmp = data.frame(symbols=allMarkers,target=target,level=targLevel)
        tmpO = data.frame(symbols=originalThree,target=target,level=targLevel)#,t(unlist(g_exp)))
        tmpO = cbind(tmpO,MedianExpression_inTarget = g_exp)
        # print(dim(tmp)[1])
        # markerSet = rbind(markerSet,tmp)
        originalMarkerSet = rbind(originalMarkerSet,tmpO)
    }
}

# Drop annotation levels we're not currently planning to use
drop_list = c("Oligodendrocytes","Microglia","Other","Astrocytes","Htr3a Vip+ - Int6","Htr3a Vip+ - Int9","Htr3a Vip+ - Int10","Htr3a Vip- Ivy NGF Int13","Htr3a Vip- Ivy NGF Int14","Htr3a Vip- Ivy NGF Int15 + 16")
targetMarkerSet = originalMarkerSet[!originalMarkerSet$target %in% drop_list,]

unique_genes = as.character(unique(targetMarkerSet$symbols))
total_unique_genes = length(unique_genes)
print(total_unique_genes)
sum(unique_genes %in% kenneths$GENE)
sum(unique_genes %in% currently_used_Genes)
kenNew = setdiff(kenneths$GENE,currently_used_Genes)
sum(unique_genes %in% kenNew)

write.csv(originalMarkerSet,file="originalMarkerSet.csv")
write.csv(targetMarkerSet,file="targetMarkerSet.csv")

median(originalMarkerSet$MedianExpression_inTarget)
min(originalMarkerSet$MedianExpression_inTarget)
ggplot(originalMarkerSet)+geom_histogram(aes(x=MedianExpression_inTarget))+scale_x_log10()
```

Find maximum median expression for a given gene over all levels of annotation. Just to help in writing the explanation.

```{r }
gene="Gad1"
maxMedian=0
for(targLevel in 1:length(levels)){
    targets = as.character(unique(levels[[targLevel]]$candidates[,1]))
    for(target in targets){
         curMax = max(levels[[targLevel]]$stats$groupMedians[gene,])
         if(curMax>maxMedian){maxMedian=curMax}
    }
}
maxMedian
```

Select how many probes to use. Requires importing how many probes we have already.

```{r }
load(file="/Users/ns9/Google Drive/In Situ Analysis/res2.rda")
colnames(res2)[1] = "symbols"
mSets = merge(targetMarkerSet,res2,by="symbols",all.x=TRUE)
mSets$nProbes[is.na(mSets$nProbes)]=0
mSets = mSets[,c(1,2,3,4,6)]
mSets$curProbes=mSets$nProbes

#gSym = as.character(unique(mSets$symbols))
minExp = aggregate(mSets$MedianExpression_inTarget,by=list(mSets$symbols),FUN=min)
colnames(minExp)=c("symbols","MedianExpression_inTarget")
nP = aggregate(mSets$nProbes,by=list(mSets$symbols),FUN=min)
colnames(nP)=c("symbols","numProbes")
combined = merge(nP,minExp,by="symbols")
combined$curProbes = combined$numProbes
combined$numProbes[combined$numProbes==0]=2

## Selective add probes where I want them
combined[combined$symbols=="Gad1","numProbes"]=6
combined[combined$symbols=="Slc6a1","numProbes"]=10

## Add new probes in a greedy manner
combined$num_probes_to_order = combined$numProbes - combined$curProbes
while(sum(combined$num_probes_to_order)<96){
    combined$multiple = combined$MedianExpression_inTarget*combined$numProbes
    combined=combined[order(combined$multiple),]
    # Restict so only five genes are ordered at most for any one gene
    addToGene=1
    for(i in 1:dim(combined)[1]){
        if(combined$num_probes_to_order[i]<5){
            addToGene=i
            break()
        }
    }
    combined$numProbes[addToGene] = combined$numProbes[addToGene]+1
    combined$num_probes_to_order = combined$numProbes - combined$curProbes
}
write.csv(combined,file="ListOfProbes.csv")

# Add the targets back
combined_wtTarg = merge(combined,targetMarkerSet[,1:2],by="symbols")
combined_wtTarg=combined_wtTarg[order(combined_wtTarg$multiple),]
write.csv(combined_wtTarg,file="ListOfProbes_with_targets.csv")
```


Generate plots of mean expression vs proportion

```{r }
kenneths = read.csv("kenneths_targets.csv")
ken_genes = unique(c(as.character(kenneths$GENE),res2$symbols))

for(targLevel in 1:length(levels)){
    targets = as.character(unique(levels[[targLevel]]$candidates[,1]))
    for(target in targets){

        # Get genes selected for this
        targGenes = as.character(originalMarkerSet[originalMarkerSet$target==target,"symbols"])
        targGenesStr = paste(targGenes,collapse=", ")
        titleStr = sprintf("%s      (%s)",target,targGenesStr)
        
        medianExp = levels[[targLevel]]$stats$groupMedians[,target]
        
        # Adjust proportions to only relate to the annotation level below
        if(targLevel>1){
            annotGroups=GetLowerAnnotGroups(target,targLevel,levels)
            annot=levels[[targLevel]]$annot
            a0=annotGroups$annotBelow        
            a0_targ = a0[annot==target][1]
            use_groups = unique(annot[a0==a0_targ])
            props = levels[[targLevel]]$stats$proportions[,use_groups]
            props2 = props/(apply(props,1,sum)+0.00000001)
            proportion = props2[,target]
        }else{
            proportion = levels[[targLevel]]$stats$proportions[,target]
        }
        
        combined = data.frame(symbols = names(proportion),proportion=proportion,median=medianExp)
        combined$isKenneths = combined$symbols  %in% ken_genes
        combined$isMine = 7.5
        combined$isMine[combined$symbols  %in% originalMarkerSet$symbols] = 10
        combined$whose = "Not Selected"
        isK=combined$symbols  %in% ken_genes
        isM=combined$symbols  %in% originalMarkerSet$symbols
        combined$whose[isM] = "Mine"
        combined$whose[isK] = "Kenneths"
        combined$whose[isK & isM] = "Both"
        
        #allGenes$isKenneths = allGenes$gene %in% kenneths$GENE
        
        pdf(file=sprintf("Mean_vs_Prop/Mean_vs_Prop---Level%s---%s.pdf",targLevel,target),width=16,height=14)
        print(ggplot(combined)+geom_text(aes(x=proportion,y=median,colour=whose,label=symbols))+scale_y_log10()+ggtitle(titleStr))
        dev.off()
    }
}

#ggplot(combined)+geom_text(aes(x=proportion_median,y=median,colour=isKenneths,label=gene))+scale_y_log10()

ggplot(combined) + geom_text(aes(x=median,y=proportion))
```